\LARGE \bf Expectation-Maximization based defense mechanism for distributed Model Predictive Control

### 2022-05-10 article/article.tex
add cheating matrix value
correct some reviewer's remarks

\emph{Notation:}
In this paper, $\norm{\cdot}$, $\norm{\cdot}_{F}$, and ${\norm{\vec{v}}_{Y}=\norm{Y^{\frac{1}{2}}\vec{v}}}$ represent the $\ell_{2}$, Frobenius and weighted norms.
$\Proj^{\set{T}}(\cdot)$ is the Euclidean projection onto $\set{T}$.
% $\card{x}$ is the number of elements in $x$.
$\kron{}{}$ ($\hadamard{}{}$) is the Kronecker (Hadamard) product.
$\vectorize{A}$ vectorizes matrix $A$.
${n\until i\until j}$ is a row vector builder with elements $\{n,n+i,\dots,n+mi\}$, where ${m=\mathrm{truncate}(\frac{j-n}{i})}$, and ${n\until j}$ means ${i=1}$.
% $[A,B]$ ($[A;B]$) is horizontal (vertical) concatenation of matrices.
${\0_{m}=0_{m\times 1}}$ (${\1_{m}=1_{m\times 1}}$), where
$0_{m\times n}$ ($1_{m\times n}$) is a $0$ ($1$) filled \mbox{$m$-by-$n$} matrix.
$\indicator{x}$ is the indicator function returning $1$ if $x$ is true and $0$ otherwise.
$\symmetric^{n}$, and $\defpos^{n}$ ($\semidefpos^{n}$) are symmetric and positive (semi-)definite matrices of size $n$.
${A^{\dagger}={{(A^{T}A)}^{-1}A^{T}}}$.
% $\determinant{A}$ is the determinant of $A$.
A vector $\vec{v}_{i}$, correspond to the $i$-th agent, and these vectors can be stacked in a vector $\vec{v}$.
% $\random{a},\random{A},\randomvec{a}$ are random variables, scalar, matrix and vector respectively.
$\expectation{\random{x}}$ is the expected value of random variable $\random{x}$, and $\probability{A \mid B}$ is the conditional probability of $A$ given condition $B$.
% $\pdf{\cdot}$ is a probability density function.
${\diag(A_{1},\dots,A_{N})}$ corresponds to a block diagonal matrix.

@@

The $P_{i}^{n}$ are constructed using $H_{i}$ and $\bar{\Gamma}_{i}$; and $\vec{s}_{i}^{j}$ are constructed using $H_{i}$, $\fik$ and $\bar{\Gamma}_{i}$. The specific construction depends on the active constraints in~\eqref{eq:constraints}.
For example, if all constraints are active, we have ${P_{i}^{\text{ac}}={(\bar{\Gamma}_{i}H_{i}^{-1}\bar{\Gamma}_{i}\T)}^{-1}}$ and ${\vec{s}_{i}^{\text{ac}}[k]=P_{i}^{\text{ac}}\bar{\Gamma}_{i}H_{i}^{-1}\fik}$.
Futhermore, if all constraints are inactive, we have ${P_{i}^{\text{in}}=0_{c\times c}}$ and ${\vec{s}_{i}^{\text{in}}=\0_{c}}$.

@@

\begin{remark}\label{rmk:P_constant}
The $\vec{s}_{i}^{j}[k]$ depend on time $k$, while $P_{i}^{j}$ do not.
\end{remark}

\begin{challenge}\label{ch:zone_unknown}
  It is important to note that since $\vec{v}_{i}$ is unknown by the coordinator, it cannot anticipate the partition of the space for each agent.
\end{challenge}
\begin{challenge}\label{ch:s_unknown}
 For the same reason, the values of $\vec{s}_{i}[k]$ are also unknown.
\end{challenge}

@@

We fix the index for these zones where all constraints are active to $1$, now referenced to as the $1$-zones.

@@

\begin{assumption}\label{ass:Pnominal}
  The nominal value of $P_{i}^{j}$ for this given zone $j$, denoted $\bar{P}_{i}^{j}$, is available from reliable attack-free historical data.
\end{assumption}

Since $P_{i}^{n}$ do not change w.r.t.\ time, we can detect a deviation from nominal behavior using ${E_{i}[k] =\|\widehat{\tilde{P}_{i}^{j}}[k]-\bar{P}_{i}^{j}\|_{F}}$.
Let ${d_{i}}$ be an indicator
\begin{equation}
  \label{eq:2}
  d_{i}=\indicator{E_{i}[k]\geq\epsilon_{P}},
\end{equation}
which detects the attack in agent $i$
if the disturbance $E_{i}[k]$ disrespects an arbitrary bound $\epsilon_{P}$.

If an attack is detected, we can estimate the inverse of $T_{i}(k)$ with
\begin{equation}
\widehat{{T_{i}(k)}^{-1}}=\bar{P}_{i}^{j}{\widehat{\tilde{P}_{i}^{j}}[k]}^{-1},
\end{equation}
if $\widehat{\tilde{P}_{i}^{j}}[k]$ is invertible, and $\widehat{\tilde{P}_{i}^{j}}[k]$ is only invertible when all constraints are active, that means, when ${j=1}$.
\begin{assumption}
  We suppose a $1$-zone exists for each agent $i$ for every time $k$.
\end{assumption}
Moreover, from~\eqref{eq:lambdafuntheta}, we can reconstruct $\vec{\lambda}_{i}$:
\begin{equation}
  \label{eq:lambda_reconstruction}
  {\vec{\lambda}_{i}}_{\mathrm{rec}}=\widehat{{T_{i}[k]}^{-1}} \tilde{\vec{\lambda}_{i}}
\end{equation}

@@

\SetKwBlock{Estep}{ E step:}{}
\SetKwBlock{Mstep}{ M step:}{}
\begin{algorithm2e}[h]
  \DontPrintSemicolon%
  Initialize parameters $\set{P}_{\mathrm{new}}$\;
  \Repeat{$\set{P}_{\mathrm{cur}}$ converges}{
    $\set{P}_{\mathrm{cur}}=\set{P}_{\mathrm{new}}$\;
    \Estep{
      Evaluate ${\probability{\set{U}|\set{B};\set{P}_{\mathrm{cur}}}}$\;
    }
    \Mstep{
      Reestimate parameters using:
      \begin{equation}
      \resizebox{.68\columnwidth}{!}{$
        \label{eq:m_step}
        \set{P}_{\mathrm{new}}=\arg\underset{\set{P}}{\max\!.}\ \expectation[{\probability{\set{U}|\set{B};\set{P}_{\mathrm{cur}}}}]{\ln\probability{\set{B},\set{U};\set{P}}}
        $}
      \end{equation}
    }
}
 \caption{Expectation Maximization}\label{alg:em}
\end{algorithm2e}

@@

As~\eqref{eq:linear_cheating} gives us a multidimensional \pwa{} function, we propose using a multidimensional expansion of the model referred to as \emph{mixture of linear regressions} in~\cite{FariaSoromenho2010} to map the relation between $\random{\Theta}$ and $\random{\Lambda}$. We will call the model \emph{mixture of affine regressions}, since our regressors have linear and constant terms:
% TODO(accacio): change subscript n
% \begin{equation}\label{eq:linear_cheating_random}
%   \randomvec{\lambda}_{o}=
%   \begin{cases}
%     -\tilde{P}^{1}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{1}&\text{if in zone } 1\\
%     \qquad\quad \vdots&\qquad\quad \vdots\\
%     -\tilde{P}^{Z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{Z}&\text{if in zone } Z\\
%   \end{cases}.
% \end{equation}
\begin{equation}\label{eq:linear_cheating_random}
  \randomvec{\lambda}_{o}=
    -\tilde{P}^{z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{z}\text{, if in zone } z,\\
\end{equation}
with  ${z\in\set{Z}=\{1\until Z\}}$.

@@

\subsection{Safe dMPC Algorithm}\label{ssec:safe_algo}
Integrating the EM to our detection and mitigation mechanism we can propose in Algorithm~\ref{alg:safeDMPC} a secure \dmpc{}, divided into two phases: detection and negotiation.
The new algorithm corresponds to adding a supervision layer for each agent as in Fig.~\ref{fig:safe_dmpc_graph}
\SetKwBlock{negotPhase}{ Negotiation Phase:}{}
\SetKwBlock{detectPhase}{ Detection Phase:}{}
\begin{algorithm2e}[h]
  \DontPrintSemicolon
  \detectPhase{
    Coordinator sends sequence of $\vec{\theta}_{i}^{o}$, $o\in\set{O}$ \;
    Subsystems solve~\eqref{eq:DOP_local}, and send $\tilde{\vec{\lambda}}^{o}_{i}$, $o\in\set{O}$\;
    Coordinator estimates $\widehat{\tilde{P}_{i}^{1}}[k]$ and $\widehat{\tilde{\vec{s}}_{i}^{1}}[k]$ with \EM{}\;
    Coordinator computes $d_{i}$ using~\eqref{eq:2}\;
  }
  \negotPhase{
  Apply Algorithm~\ref{alg:quantityAlg} with adequate versions of $\vec{\lambda}_{i}\p$:\;
  \quad$\tilde{\vec{\lambda}}_{i}\p$, if $d_{i}=0$ and ${\vec{\lambda}_{i}}_{\mathrm{rec}}$, if ${d_{i}=1}$~\eqref{eq:lambda_reconstruction}\;
 }
 \caption{Secure DMPC.}\label{alg:safeDMPC}
\end{algorithm2e}

\section{Example: Temperature Control}\label{sec:App_dMPC}
The system consists of $4$~distinct~rooms (I to IV), which we want to control the air temperatures inside each one of them.
The systems are modeled as continuous-time linear \mbox{time-invariant} systems using the \mbox{3R-2C} model, with parameters in Tables~\ref{tab:modelParamMeaning} and~\ref{tab:modelParam}, and dynamics
\begin{equation}
\begin{matrix}
  \label{eq:systems_cont}
\dot{\vec{x}}_{i}(t)&=&{A_{c}}_{i}\vec{x}_{i}(t) &+& {B_{c}}_{i}\vec{u}_{i}(t)\\
\vec{y}_{i}(t)&=&{C_{c}}_{i}\vec{x}_{i}(t) &&
\end{matrix},
\end{equation}
with
\begin{equation*}
  \label{eq:room_system}
  \begin{matrix}
\resizebox{.98\columnwidth}{!}{$
  A_{\mathrm{c}_{i}}=\left[
    \begin{matrix}
      -\frac{1}{C^{\text{walls}}_{i}R^{\text{oa/ia}}_{i}}-\frac{1}{C^{\text{walls}}_{i}R^{\text{iw/ia}}_{i}}& \frac{1}{C^{\text{walls}}_{i}R^{\text{iw/ia}}_{i}}\\
      \frac{1}{C^{\text{air}}_{i}R^{\text{iw/ia}}_{i}} &-\frac{1}{C^{\text{air}}_{i}Ro_{i}}-\frac{1}{C^{\text{air}}_{i}R^{\text{iw/ia}}_{i}}
    \end{matrix}\right]
    $}
  \\
  \begin{matrix}
    B_{\mathrm{c}_{i}}=\left[
      \begin{matrix}  \frac{10}{C^{\text{walls}}_{i}}& 0\end{matrix}
    \right]\T&C_{\mathrm{c}_{i}}=\left[\begin{matrix}1 & 0\end{matrix}\right]
  \end{matrix}
  \end{matrix},
\end{equation*}
where ${\vec{x}_{i}=[{{x}_{A}}_{i}\T\ {{x}_{W}}_{i}\T]\T}$. ${x_A}_i$ and ${x_W}_i$ are the mean temperatures of the air and walls inside room~$i$. $\vec{u}_{i}$ is the input (the heating power)
for the corresponding room. The inputs are constraint by ${\sum_{i=1}^{4}\vec{u}_{i}(t)\preceq 4\mathrm{kW}}$.

@@

>>>>

\emph{Notation:}
In this paper, $\norm{\cdot}$, $\norm{\cdot}_{F}$, and ${\norm{\vec{v}}_{Y}=\norm{Y^{\frac{1}{2}}\vec{v}}}$ represent the $\ell_{2}$, Frobenius and weighted norms.
$\Proj^{\set{T}}(\cdot)$ is the Euclidean projection onto $\set{T}$.
% $\card{x}$ is the number of elements in $x$.
$\kron{}{}$ ($\hadamard{}{}$) is the Kronecker (Hadamard) product.
$\vectorize{A}$ vectorizes matrix $A$.
${n\until i\until j}$ is a row vector builder with elements $\{n,n+i,\dots,n+mi\}$, where ${m=\mathrm{truncate}(\frac{j-n}{i})}$, and ${n\until j}$ means ${i=1}$.
% $[A,B]$ ($[A;B]$) is horizontal (vertical) concatenation of matrices.
${\0_{m}=0_{m\times 1}}$ (${\1_{m}=1_{m\times 1}}$), where
$0_{m\times n}$ ($1_{m\times n}$) is a $0$ ($1$) filled \mbox{$m$-by-$n$} matrix.
$\indicator{x}$ is the indicator function returning $1$ if $x$ is true and $0$ otherwise.
$\symmetric^{n}$, and $\defpos^{n}$ ($\semidefpos^{n}$) are symmetric and positive (semi-)definite matrices of size $n$.
${A^{\dagger}={{(A^{T}A)}^{-1}A^{T}}}$.
% $\determinant{A}$ is the determinant of $A$.
A vector $\vec{v}_{i}$, corresponds to the $i$-th agent, and these vectors can be stacked in a vector $\vec{v}$.
% $\random{a},\random{A},\randomvec{a}$ are random variables, scalar, matrix and vector respectively.
$\expectation{\random{x}}$ is the expected value of random variable $\random{x}$, and $\probability{A \mid B}$ is the conditional probability of $A$ given condition $B$.
% $\pdf{\cdot}$ is a probability density function.
${\diag(A_{1},\dots,A_{N})}$ corresponds to a block diagonal matrix.

@@

The $P_{i}^{n}$ are constructed using $H_{i}$ and $\bar{\Gamma}_{i}$; and $\vec{s}_{i}^{n}$ are constructed using $H_{i}$, $\fik$ and $\bar{\Gamma}_{i}$. The specific construction depends on the active constraints in~\eqref{eq:constraints}.
For example, if all constraints are active, we have ${P_{i}^{\text{ac}}={(\bar{\Gamma}_{i}H_{i}^{-1}\bar{\Gamma}_{i}\T)}^{-1}}$ and ${\vec{s}_{i}^{\text{ac}}[k]=P_{i}^{\text{ac}}\bar{\Gamma}_{i}H_{i}^{-1}\fik}$.
Futhermore, if all constraints are inactive, we have ${P_{i}^{\text{in}}=0_{c\times c}}$ and ${\vec{s}_{i}^{\text{in}}=\0_{c}}$.

@@

\begin{remark}\label{rmk:P_constant}
The $\vec{s}_{i}^{n}[k]$ depend on time $k$, while $P_{i}^{n}$ do not.
\end{remark}

\begin{challenge}\label{ch:zone_unknown}
  It is important to note that since $\vec{v}_{i}[k]$ is unknown by the coordinator, it cannot anticipate the partition of the space for each agent.
\end{challenge}
\begin{challenge}\label{ch:s_unknown}
 For the same reason, the values of $\vec{s}_{i}[k]$ are also unknown.
\end{challenge}

@@

We fix the index for the zones where all constraints are active to $1$, now referenced to as zones $1$ or the $1$-zones.

@@

\begin{assumption}\label{ass:Pnominal}
  The nominal value of $P_{i}^{j}$ for this given $j$-zone, denoted $\bar{P}_{i}^{j}$, is available from reliable attack-free historical data.
\end{assumption}

Since $P_{i}^{j}$ do not change w.r.t.\ time, we can detect a deviation from nominal behavior using ${E_{i}[k] =\|\widehat{\tilde{P}_{i}^{j}}[k]-\bar{P}_{i}^{j}\|_{F}}$.
Let ${\mathfrak{D}_{i}}$ be an indicator
\begin{equation}
  \label{eq:2}
  \mathfrak{D}_{i}=\indicator{E_{i}[k]\geq\epsilon_{P}},
\end{equation}
which detects the attack in agent $i$
if the disturbance $E_{i}[k]$ disrespects an arbitrary bound $\epsilon_{P}$.

If an attack is detected, we can estimate the inverse of $T_{i}(k)$ with
\begin{equation}
\widehat{{T_{i}(k)}^{-1}}=\bar{P}_{i}^{j}{\widehat{\tilde{P}_{i}^{j}}[k]}^{-1},
\end{equation}
if $\widehat{\tilde{P}_{i}^{j}}[k]$ is invertible, and $\widehat{\tilde{P}_{i}^{j}}[k]$ is only invertible when all constraints are active, that means, when ${j=1}$.
\begin{assumption}
  For each agent $i$, at every time $k$, there is a corresponding $1$-zone.
\end{assumption}
Moreover, from~\eqref{eq:lambdafuntheta}, we can reconstruct $\vec{\lambda}_{i}$:
\begin{equation}
  \label{eq:lambda_reconstruction}
  {\vec{\lambda}_{i}}_{\mathrm{rec}}=\widehat{{T_{i}[k]}^{-1}} \tilde{\vec{\lambda}_{i}}
\end{equation}

@@

\SetKwBlock{Estep}{ E step:}{}
\SetKwBlock{Mstep}{ M step:}{}
\begin{algorithm2e}[h]
  \DontPrintSemicolon%
  Initialize parameters $\set{P}_{\mathrm{new}}$\;
  \Repeat{$\set{P}_{\mathrm{cur}}$ converges}{
    $\set{P}_{\mathrm{cur}}\gets\set{P}_{\mathrm{new}}$\;
    \Estep{
      Evaluate ${\probability{\set{U}|\set{B};\set{P}_{\mathrm{cur}}}}$\;
    }
    \Mstep{
      Reestimate parameters using:
      \begin{equation}
      \resizebox{.68\columnwidth}{!}{$
        \label{eq:m_step}
        \set{P}_{\mathrm{new}}=\arg\underset{\set{P}}{\max\!.}\ \expectation[{\probability{\set{U}|\set{B};\set{P}_{\mathrm{cur}}}}]{\ln\probability{\set{B},\set{U};\set{P}}}
        $}
      \end{equation}
    }
}
 \caption{Expectation Maximization}\label{alg:em}
\end{algorithm2e}

@@

As~\eqref{eq:linear_cheating} gives us a multidimensional \pwa{} function, we propose using a multidimensional expansion of the model referred to as \emph{mixture of linear regressions} in~\cite{FariaSoromenho2010} to map the relation between $\random{\Theta}$ and $\random{\Lambda}$. We will call the model \emph{mixture of affine regressions}, since our regressors have linear and constant terms:
% TODO(accacio): change subscript n
% \begin{equation}\label{eq:linear_cheating_random}
%   \randomvec{\lambda}_{o}=
%   \begin{cases}
%     -\tilde{P}^{1}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{1}&\text{if in zone } 1\\
%     \qquad\quad \vdots&\qquad\quad \vdots\\
%     -\tilde{P}^{Z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{Z}&\text{if in zone } Z\\
%   \end{cases}.
% \end{equation}
\begin{equation}\label{eq:linear_cheating_random}
  \randomvec{\lambda}_{o}=
    -\tilde{P}^{z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{z}\text{, if in zone } z,\\
\end{equation}
with  ${z\in\set{Z}=\{1\until Z\}}$.
\remark{Observe that the indices $z$, do not necessarily correspond to the original indices $n$.}

@@

\subsection{Safe dMPC Algorithm}\label{ssec:safe_algo}
Integrating the EM to our detection and mitigation mechanism we can propose in Algorithm~\ref{alg:safeDMPC} a secure \dmpc{}, divided into two phases: detection and negotiation.
The new algorithm corresponds to adding a supervision layer for each agent as in Fig.~\ref{fig:safe_dmpc_graph}
\SetKwBlock{negotPhase}{ Negotiation Phase:}{}
\SetKwBlock{detectPhase}{ Detection Phase:}{}
\begin{algorithm2e}[h]
  \DontPrintSemicolon
  \detectPhase{
    Coordinator sends sequence of $\vec{\theta}_{i}^{o}$, $o\in\set{O}$ \;
    Subsystems solve~\eqref{eq:DOP_local}, and send $\tilde{\vec{\lambda}}^{o}_{i}$, $o\in\set{O}$\;
    Coordinator estimates $\widehat{\tilde{P}_{i}^{1}}[k]$ and $\widehat{\tilde{\vec{s}}_{i}^{1}}[k]$ with \EM{}\;
    Coordinator computes $\mathfrak{D}_{i}$ using~\eqref{eq:2}\;
  }
  \negotPhase{
  Apply Algorithm~\ref{alg:quantityAlg} with adequate versions of $\vec{\lambda}_{i}\p$:\;
  \quad$\tilde{\vec{\lambda}}_{i}\p$, if $\mathfrak{D}_{i}=0$ and ${\vec{\lambda}_{i}}_{\mathrm{rec}}$, if ${\mathfrak{D}_{i}=1}$~\eqref{eq:lambda_reconstruction}\;
 }
 \caption{Secure DMPC.}\label{alg:safeDMPC}
\end{algorithm2e}

\section{Example: Temperature Control}\label{sec:App_dMPC}
The system consists of $4$~distinct~rooms (I to IV), which we want to control the air temperatures inside each one of them.
The systems are modeled as continuous-time linear \mbox{time-invariant} systems using the \mbox{3R-2C} model, with parameters in Tables~\ref{tab:modelParamMeaning} and~\ref{tab:modelParam}, and dynamics
\begin{equation}
\begin{matrix}
  \label{eq:systems_cont}
\dot{\vec{x}}_{i}(t)&=&{A_{c}}_{i}\vec{x}_{i}(t) &+& {B_{c}}_{i}\vec{u}_{i}(t)\\
\vec{y}_{i}(t)&=&{C_{c}}_{i}\vec{x}_{i}(t) &&
\end{matrix},
\end{equation}
with
\begin{equation*}
  \label{eq:room_system}
  \begin{matrix}
\resizebox{.98\columnwidth}{!}{$
  A_{\mathrm{c}_{i}}=\left[
    \begin{matrix}
      -\frac{1}{C^{\text{walls}}_{i}R^{\text{oa/ia}}_{i}}-\frac{1}{C^{\text{walls}}_{i}R^{\text{iw/ia}}_{i}}& \frac{1}{C^{\text{walls}}_{i}R^{\text{iw/ia}}_{i}}\\
      \frac{1}{C^{\text{air}}_{i}R^{\text{iw/ia}}_{i}} &-\frac{1}{C^{\text{air}}_{i}Ro_{i}}-\frac{1}{C^{\text{air}}_{i}R^{\text{iw/ia}}_{i}}
    \end{matrix}\right]
    $}
  \\
  \begin{matrix}
    B_{\mathrm{c}_{i}}=\left[
      \begin{matrix}  \frac{10}{C^{\text{walls}}_{i}}& 0\end{matrix}
    \right]\T&C_{\mathrm{c}_{i}}=\left[\begin{matrix}1 & 0\end{matrix}\right]
  \end{matrix}
  \end{matrix},
\end{equation*}
where ${\vec{x}_{i}=[{{x}_{A}}_{i}\T\ {{x}_{W}}_{i}\T]\T}$. ${x_A}_i$ and ${x_W}_i$ are the mean temperatures of the air and walls inside room~$i$. $\vec{u}_{i}$ is the input (the heating power)
for the corresponding room. The inputs are constrained by ${\sum_{i=1}^{4}\vec{u}_{i}(t)\preceq 4\mathrm{kW}}$.

@@

  For scenarios (2) and (3), Agent I uses
  $$T_{I}=\left[\input{triche_matrix}\right].$$
